---
title: "BOO 2025 - Example Analysis"
subtitle: "Script 1: Data exploration - Questions"
date: "`r Sys.Date()`" 
author: 
  Pien van Zon
output:
  html_document:
    code_download: true
    theme: united
    highlight: tango
    df_print: paged
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

***

> In this script you will explore the project data, describing the variables and outlining the data structure. 

***

# Setup

## Using this template

For each of the steps in the BOO project, we have prepared an example script like this one. You can download the R markdown (.Rmd) file to make it easier to complete the steps.

**Exercise 1: Download the .Rmd using the `Code` dropdown menu in the top right corner of this document, and load it into R Studio.**

***

## Clean

It is good practice to clean your environment at the start of every script. That way all the objects you need for the analysis are loaded within the code, and the R markdown document will knit without issues. 

**Exercise 2: Remove all current objects from the R environment:**

<details>
  <summary><strong>Hint</strong></summary>

  `ls()` is a function that returns the names of all objects currently loaded into the session.

</details>

```{r clean}
# This clears the environment.
rm(list=ls())

```

***

## Load packages

In R, different packages can have functions with the same name. Later loaded packages can therefore mask functions from earlier loaded ones. To avoid confusion when rerunning chunks within a script, we advise loading all required packages at the start. 

As you learned in the DataCamp course, `tidyverse` is a group of packages which includes many basic functions for data wrangling and visualization in R.

**Exercise 3: Load the `tidyverse` package into your environment:**

```{r load-packages}
# The tidyverse is already loaded.
library(tidyverse)

```

You can see from the reported message that loading `tidyverse` masks base R functions, `filter()` and `lag()`. If it is the case that a function you want to use is masked, you can always specify the package you intend to draw from explicitly using the `::` operator (e.g. `stats::filter()`).

***

## Set variables

Setting variables that you will use repeatedly in a script can save time if you need to make changes later. This can include paths to folders where you store files or QC thresholds, for example.

**Exercise 4: Create the following objects in your R environment:**

* `root_dir` - your project folder
* `count_path` - location of the count data
* `metadata_path` - location of the metadata (sample information)

```{r set-variables}
# the values are now in my environment
root_dir <- "C:/Users/pienz/Universiteit Leiden/BOO 2025 - BOO CDS Giulia team - BOO CDS Giulia team/Students/Pien/project/pienvanzon_BOO/"
count_path <- file.path(root_dir, "input", "countData.txt")
metadata_path<- file.path(root_dir, "input", "metaData.txt")

```

***

## Load data

**Exercise 5: Load the project data into your R environment:**

<details>
  <summary><strong>Hint</strong></summary>

  Use the `read_tsv()` function.

</details>

```{r load-data}
# This reads the count data
countData <- read_tsv(count_path)

# This shows al the count data
str(countData)

#This reads the data from my file
metaData <- read_tsv(metadata_path)

#This shows the metaData
str(metaData)
```

countData:
- contains one column containing probe names
- The other columns are samples that contain counts of probes that are detected in a sample
- this has one row for each probe measured by the TempO-Seq

metaData:
- It should have a row for each of the sample columns in countData
- It should have variables describing the samples in your data in the columns

***

# Metadata

The `metaData` object contains information on the individual samples in the data.

**Exercise 6: Briefly describe each of the variables in `metaData`:**

```{r metadata}
# This gives the different names of the columns
colnames(metaData)

# This gives the names of what comes under the columns.
head(metaData)

```
- sample_ID is the ID for the sample (same as the column names of the countData).
- sample_type tells us if the sample is a Treatment or Control sample
- compound_ID is the ID of the two compounds in this data (CS1119 and CS1131)
- compound_CAS is the CAS numbers for the two compounds
- compound_name is the name of the compounds
- compound_abbr is a shorter name for the compounds (for plots)
- mean_ID is the group ID composed of the compound_ID, conc_ID, and plate number
- mean_ID_control is the mean_ID of the control samples that this sample should be compared to
- control_type is the type of control that should be used (Medium, DMSOLow, or DMSOHigh)
- conc_ID is the ID of the concentration
- conc_amt is the amount of concentration
- conc_unit is the unit of the concentration
- conc_max is the maximum concentration for this compound
- conc_min is the minimum concentration for this compound
- rep is the replicate number
- lib_size is the library size (total number of reads profiled from this sample)
- read_depth is the average read depth (library size / number of probes profiled)
- plate_ID is the plate number
- plate_well is the location of the sample on the plate
- plate_row is the row of the plate this sample was on
- plate_col is the column of the plate this sample was in
- batch is the batch number
- timepoint is the number of hours after exposure that expression was profiled
- cell_type is the cell type (PHH for all samples)
- compound_class is High or Low toxicity (or NA for controls)
- LOAEL_amt is the lowest observed adverse effect level (LOAEL)
- LOAEL_unit is the unit of the LOAEL (mg/kg/d)
- PBPK_equivalent is the in vitro equivalent of PBPK_serum level (10mg/kg)



***

**Exercise 7: What test system is this data profiled from? Give a brief description.**

<details>
  <summary><strong>Hint</strong></summary>

  Try looking at the `cell_type` variable.

</details>

```{r test-system}
# This function gives the cell type of the metaData which gives PHH
unique(metaData$cell_type)


```

***

**Exercise 8: How many treatment and control samples are there in the project data?**

```{r sample-type}
# This gives the output of the control and treatment. There are 18 controls and 42 treatments.
table(metaData$sample_type)

```

***

**Exercise 9: Describe the range of experimental conditions which the samples have been exposed to:**

<details>
  <summary><strong>Hint</strong></summary>

  Don't forget to describe the compounds and the concentration ranges.

</details>

```{r experimental-conditions}
# 
unique(metaData$compound_name)
table(metaData$compound_name, metaData$compound_class)

```
```{r}
# Show distinct combinations of compound and concentration variables
metaData %>%
  select(compound_name, conc_amt, conc_unit, conc_ID, compound_class) %>%
  distinct()
  
```


***

**Exercise 10: How many replicates are there for each experimental condition?**

```{r replicate-number}
# In the last column replicates are shown. For example the compound 4-chloro-3-methylphenol has 3 replicates with a concentration of 0,03 µM.
metaData %>%
  group_by(compound_name, conc_amt, conc_unit, conc_ID, compound_class) %>%
  summarise(n_replicates = n())

```

***

**Exercise 11: Are the compounds on different plates? If so, how might this affect the analysis and results? Has the experimental design been adjusted to compensate for this?**

```{r plates}
# The table shows that the 4-chloro-3-methylphenol on P6 is plated, just as its control DMSO. Laurylamine dipropylenediamine is plated on P4, just like its control, Medium.
table(metaData$compound_name, metaData$plate_ID)

```

***

**Exercise 12: Why do some samples not have data on their toxicity class?**

<details>
  <summary><strong>Hint</strong></summary>

  You can filter data using the `is.na()` function.

</details>

```{r na-class}
# The toxicity class is missing in the DMSO and Mdeium compounds, this is because they are not toxic. These are the controls and thus not relevant.
metaData_missing <- metaData %>% filter(is.na(compound_class))
metaData_missing
```

***

**Exercise 13: How many hours after exposure are the samples collected?**

```{r timepoint}
# The samples are collected after 24 hours of exposure. This is done 60 times.
table(metaData$timepoint)

```

***

**Exercise 14: Check that each sample has a unique identifier:**

```{r unique-id}
# Check if the length of unique sample IDs matches the number of rows in the metadata
length(unique(metaData$sample_ID)) == nrow(metaData)

```
Each sample has a unique identifier, located in the sample_ID column.

***

# Count data

The count data has information on raw counts from the probes measured during TempO-Seq (rows) in each of the samples (columns)

Many of the functions we will use in this project assume that `countData` row names are the gene or probe names. 

**Exercise 15: Change the variable `Probe` to the row names:**

<details>
  <summary><strong>Hint</strong></summary>

  Use the `column_to_rownames()` function.

</details>

```{r column-rownames}
# The table has column and row names now.
countData <- countData %>% 
  column_to_rownames("Probe")
```

***

**Exercise 16: Which column in the metadata can be used to subset the count data? Use it to select only counts from the control samples:**

<details>
  <summary><strong>Hint</strong></summary>

  Make sure you don't replace `countData` with only control samples when answering this question! :)

</details>

```{r}
# Check if the column names in count data are the same as the sample ID's in the meta data set
table(colnames(countData) %in% metaData$sample_ID)
```


```{r subset-controls}
# Save a vector of sample IDs for the control samples
control_IDs <- (metaData %>% filter(sample_type == "Control"))$sample_ID

# Subset the count data for only these control samples (showing only the first 100 rows)
countData[1:100, control_IDs]
```

***

**Exercise 17: Check that the columns of the count data and rows of the metadata are in the same order.**

<details>
  <summary><strong>Hint</strong></summary>

  The different functionalities of `==` and `%in%` may help here.

</details>

```{r check-order}
# check if the column names of the count data are identical to the sample ID column
table(colnames(countData) == metaData$sample_ID)
```
The order of these columns is correct.

***

# Advanced question

This question provides a bit of a (hopefully fun) challenge, but you can also skip it, save your data, and move on to the probe QC if you prefer.

**Exercise 18: Make a plot of the plate layout using `ggplot`?**

<details>
  <summary><strong>Hint</strong></summary>

  Use `geom_tile()`

</details>

```{r plate-layout}
# 

```

***

# Save

Make sure you save any changed objects at the end of your script for downstream analysis.

**Exercise 19: Save your `countData` and `metaData` as `.Rdata` objects**

<details>
  <summary><strong>Hint</strong></summary>

  Don't forget to use `root_dir`, so if you ever move your project your code can be easily updated.

</details>

```{r save-data}
# 
save(countData, file=file.path(root_dir, "output/countData_01exploration.RData"))
save(metaData, file=file.path(root_dir, "output/metaData_01exploration.RData"))

```

***

# Session Info

Printing `session_info()` at the bottom of your scripts ensures your work is reproducible. It does this by keeping a record of the package versions used and the environment. 

**Exercise 20: Print your session info here:**

```{r session-info}
# 

```

***

**Exercise 21: Knit this script to produce either a `PDF` or `HTML`file and push it to your GitHub repository.**

That is the end of the Data Exploration. Example answers will be available from the `BOO_template` GitHub on Tuesday. 

Next, please move on to the probe QC using `02_ProbeQC_Outline.Rmd`.

***

